---
title: "3: Coding Basics"
author: "Environmental Data Analytics | Kateri Salk"
date: "Spring 2020"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Objectives
1. Discuss and navigate different data types in R
2. Create, manipulate, and explore datasets
3. Call packages in R

## Data Types in R
R treats objects differently based on their characteristics. For more information, please see: https://www.statmethods.net/input/datatypes.html. 

* **Vectors** 1 dimensional structure that contains elements of the same type.

* **Matrices** 2 dimensional structure that contains elements of the same type.

* **Arrays** Similar to matrices, but can have more than 2 dimensions. We will not delve into arrays in depth.

* **Lists** Ordered collection of elements that can have different modes.

* **Data Frames** 2 dimensional structure that is more general than a matrix. Columns can have different modes (e.g., numeric and factor). When we import csv files into the R workspace, they will enter as data frames.

Define what each new piece of syntax does below (i.e., fill in blank comments). Note that the R chunk has been divided into sections (# at beginning of line, ---- at end)
```{r}
# Vectors ----
vector1 <- c(1,2,5.3,6,-2,4) # numeric vector
vector1
vector2 <- c("one","two","three") # character vector
vector2
vector3 <- c(TRUE,TRUE,TRUE,FALSE,TRUE,FALSE) #logical vector
vector3

vector1[3] 
#Line 39 is saying that the third item is 5.3. [] allow you to call up the 3rd item
# "c" means concatenate, i.e. takes everything in parenthesis and names it. entering vector1 will allow you to call up the data points after you name it.
# have to put parenthesis around specific character strings or R will think you have to do an operation called one. 

# Matrices ----
#going to 2 dimensions. The ones below are functions because R knows that the word matrix means make a matrix
matrix1 <- matrix(1:20, nrow = 5,ncol = 4) # has numbers 1-20 in it, with 5 rows and 4 columns
matrix1
matrix2 <- matrix(1:20, nrow = 5, ncol = 4, byrow = TRUE) #"byrow" means responses are by rows instead of columns
matrix2
matrix3 <- matrix(1:20, nrow = 5, ncol = 4, byrow = TRUE, # return after comma continues the line
                  dimnames = list(c("uno", "dos", "tres", "cuatro", "cinco"), 
                                  c("un", "deux", "trois", "cat"))) #dimnames names rows by whatever you want them to be called. this is calls the columns by french numbers 1-4 and the columns spanish numbers 1-4 

matrix1[4, ] # calls up matrix 1, row 4
matrix1[ , 3] # calls up matrix 1, column 3
matrix1[c(12, 14)] # calls up matrix 1, numbers 12 and 14. specifying specific objects within matrix
matrix1[c(12:14)] # calls up matrix 1, numbers 12 through 14. specifying specific objects within matrix
matrix1[2:4, 1:3] # calls up matrix 1, columns 2 to 4 by rows 1 to 3
matrix3
matrix3[c(12,14)] # get 7 and 15 because in positions 12 and 14, these are the numbers in those positions

cells <- c(1, 26, 24, 68)
rnames <- c("R1", "R2")
cnames <- c("C1", "C2") 
matrix4 <- matrix(cells, nrow = 2, ncol = 2, byrow = TRUE,
  dimnames = list(rnames, cnames)) # 
matrix4
#allows you to build a matrix from scratch. line 60, call up cells, a character vector named rnames and one named cnames
# values that will go into the matrix are "cells" by 2 rows and 2 columns, and numbers arranged across rows
# add the names by including "dimnames = list(name1, name2)"

# Lists ---- 
list1 <- list(name = "Maria", mynumbers = vector1, mymatrix = matrix1, age = 5.3); list1
list1[[2]]

# list has a name called maria, a numeric vector called vector1, a matrix, and an age
# putting in a semicolon allows you to combine the two codes so you only have to run 1 line of code
# this list function lists out the separate elements of the same list
# line 74 calls up the 2nd element in list 1
# pulling from the vectors and matrices we made above. have to make those first to do the list

# Data Frames ----
d <- c(1, 2, 3, 4) # What type of vector? numerical
e <- c("red", "white", "red", NA) # What type of vector? character
f <- c(TRUE, TRUE, TRUE, FALSE) # What type of vector? logical
dataframe1 <- data.frame(d,e,f) # makes a data frame with the 3 objects called dataframe1
# vectors show up in values section of environment and the matrices show up in data
# R recognizes NA as its own thing, so don't need to put it in quotes
# turns character vector into factor-type of data frame. in Env says "factor w/2 levels...1 2 1 NA"
names(dataframe1) <- c("ID","Color","Passed"); View(dataframe1) # can rename column titles using this format

dataframe1[1:2,] # 2 dimensions, but each column has a different mode included in single data frame
# matric subsetting is always row, columns
dataframe1[c("ID","Passed")] # calls up specific elements of data frame, i.e. "ID" and "Passed"
dataframe1$ID # Just calls up individual column

```
Question: How do the different types of data appear in the Environment tab?

> Answer: Data frames, lists, and matrices are located in the Data tabs, whereas, the objects like cells, names of rows and columns, and vectors are under the Values tab. 

Question: In the R chunk below, write "dataframe1$". Press `tab` after you type the dollar sign. What happens?

> Answer: It lists the name of the columns in dataframe1 and pushing enter calls up the last object that was called up before

### Coding challenge

Find a ten-day forecast of temperatures (Fahrenheit) for Durham, North Carolina. Create two vectors, one representing the high temperature on each of the ten days and one representing the low.

```{r}
vectorhighf <- c (62, 45, 50, 50, 40,40, 45, 52,54, 58)
vectorhighf
vectorlowf <- c(31, 29, 43, 24, 23, 23, 26, 30, 42, 41)
vectorlowf
```

Now, create two additional vectors that include the ten-day forecast for the high and low temperatures in Celsius.

```{r}
vectorhighc <- c(17, 7, 10, 10, 4, 4, 7, 11, 12, 14) # numeric vector
vectorhighc
vectorlowc <- c(-1, -1, 6, -4, -4, -5, -3, -1, 5, 5) 
vectorlowc
vectorhighc*(-32*(5/9))

```

Combine your four vectors into a data frame and add informative column names.

```{r}

dataframetemps <- data.frame(vectorhighf, vectorlowf, vectorhighc, vectorlowc) 
names(dataframetemps) <- c("High Temp F","Low Temp F","High Temp C", "Low Temp C"); View(dataframetemps) 

dataframetemps[1:2,] 
dataframetemps[c("High Temp F","High Temp C")] # calls up specific elements of data frame, i.e. "ID" and "Passed"
dataframetemps$`High Temp F`
```

Use the common functions `summary` and `sd` to obtain basic data summaries of the ten-day forecast. How would you call these functions differently for the entire data frame vs. a single column? Attempt to demonstrate both options below.

```{r}
summary.data.frame(dataframetemps)
```


## Packages

The Packages tab in the notebook stores the packages that you have saved in your system. A checkmark next to each package indicates whether the package has been loaded into your current R session. Given that R is an open source software, users can create packages that have specific functionalities, with complicated code "packaged" into a simple commands.

If you want to use a specific package that is not in your libaray already, you need to install it. You can do this in two ways: 

1. Click the install button in the packages tab. Type the package name, which should autocomplete below (case matters). Make sure to check "intall dependencies," which will also install packages that your new package uses. 

2. Type `install.packages("packagename")` into your R chunk or console. It will then appear in your packages list. You only need to do this once. 

If a package is already installed, you will need to load it every session. You can do this in two ways: 

1. Click the box next to the package name in the Packages tab. 

2. Type `library(packagename)` into your R chunk or console.


```{r}

# comment out install commands, use only when needed and re-comment

library(dplyr)
library(ggplot2)



# Some packages are umbrellas under which other packages are loaded

library(tidyverse)
```

Question: What happens in the console when you load a package?

> Answer: 


## Tips and Tricks

* Sequential section headers can be created by using at least four -, =, and # characters.

* The command `require(packagename)` will also load a package, but it will not give any error or warning messages if there is an issue.

* You may be asked to restart R when installing or updating packages. Feel free to say no, as this will obviously slow your progress. However, if the functionality of your new package isn't working properly, try restarting R as a first step. 

* If asked "Do you want to install from sources the packages which needs compilation?", type `yes` into the console. 

* You should only install packages once on your machine. If you store `install.packages` in your R chunks/scripts, comment these lines out. 

* Update your packages regularly! 